---
title: "Pre-course preparation"
author: "Denis Rustand (denis.rustand@u-bordeaux.fr)"
date: "July 8 and 9, 2026"
output:
  pdf_document: default
  html_document:
    df_print: paged
subtitle: |
  Bayesian survival, longitudinal and joint models with INLA \
  VIBASS 9, Valencia, Spain
---

# Before the course

Running the code during the course is **optional**. If you want to run the code, please complete the following steps **before the course**. Installing packages can take time, so do not leave this for the last minute.

## R version

You need **R \>= 4.6**. Check with:

```{r check-r, eval=FALSE}
R.version.string
```

If your version is older, update from <https://cloud.r-project.org>.

## Package installation

```{r install, eval=FALSE}
# Bioconductor packages required by INLA
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install(c("graph", "Rgraphviz"), dep = TRUE)

# INLA (not on CRAN)
install.packages("INLA",
  repos = c(getOption("repos"),
            INLA = "https://inla.r-inla-download.org/R/testing"),
  dep = TRUE)

# INLAjoint
install.packages("INLAjoint")

# Other packages
install.packages(c("devtools", "ggplot2", "ggpubr", "survival", 
                   "splines", "JMbayes2", "mvtnorm", "PermAlgo", 
                   "eha", "rnaturalearth", "rnaturalearthdata",
                   "spdep", "sf", "Matrix", "fmesher",
                   "terra", "tidyterra"))
```

## Verify installation

Run the following to check that the core packages load and compute correctly:

```{r verify, eval=FALSE}
library(INLA)
library(INLAjoint)
cat("INLA version:", inla.version()$version, "\n")
cat("INLAjoint version:", as.character(packageVersion("INLAjoint")), "\n")

# Toy example: intercept of y ~ 1 with two observations (2 and 8) should be 5
m <- inla(y ~ 1, data = list(y = c(2, 8)))
cat("Intercept (expected 5):", round(m$summary.fixed["(Intercept)", "mean"], 1), "\n")
```

On Linux, if INLA throws a binary error, try switching binaries:

```{r linux-fix, eval=FALSE}
inla.binary.install()
```

## Common issues

- **Linux users**: you may need system libraries `libudunits2-dev`, `libgdal-dev`, `libgeos-dev`, `libproj-dev`, `libsqlite3-dev` and `libgsl-dev`.
- **Mac users**: make sure Xcode command line tools are installed (`xcode-select --install`)
- **Windows users**: Rtools may be required for some packages
